Likelihoods for a general class of ARGs under the SMC

### 2025-04-29 paper.tex
Updates

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
    \caption{Log likelihood-ratio curves for ARGs with various properties.
    The ``Full ARG'' here is the result of an
    msprime simulation of the SMC (100 diploid samples, genome length 1Mb,
    $r=\mu=2.5\times10^{-8}$, $N_e=1\times10^4$).
    The ``RE nodes removed'' ARG is the
    result of simplifying out all of the recombination nodes,
    and the ``Fully simplified'' ARG has all locally unary nodes removed.
    Each of these simulated ARGs additionally has 5\% and 25\% of the
    internal nodes removed to create polytomies.
    We also show the results for ARGs inferred from the original simulation
    data using tsinfer and tsdate
    (see text for details), in which we use the default output of tsinfer
    including unary nodes (tsinfer+tsdate), and in which we fully simplify
    before passing to tsdate (tsinfer+tsdate FS).
    See the text for details on how the likelihood ratio
    values are computed.
    The ``Full ARG'' and ``RE nodes removed'' lines coincide.
    \label{fig:lik-surface}}
\end{figure}

@@

Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
and even on the fully simplified
ARG has only a very minor effect on location of the minimum.
Supplemental Fig.~S1 explores more extreme examples in which we remove
25\%, 50\% and 75\% of the internal nodes in these ARGs, and demonstrates 
a remarkable robustness to the loss of the majority of the internal
nodes in these examples. Interestingly, the loss of 75\% of the 
internal nodes  in the fully simplified ARG 
(1841 of 2455, leading  to an average of 27 polytomies per tree)
appears to improve performance, substantially 
reducing the bias away from the true parameter value.

Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from the  \revpoint{2}{35}
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the
\cite{li2003modeling} copying process. Information about
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to
attempt to reconstruct the exact subtree-prune-and-regraft
moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well \revpoint{2}{37}
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we
fully simplify the output of tsinfer
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a bias \revpoint{2}{38}
away from the true value of the recombination rate parameter.

\subsection*{Piecewise decomposition over time}
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under the Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices:
    from 0 to 1,000 generations ago $N_e$ is 10,000;
    from 1,000 to 5,000 generations ago it is 5,000;
    and past that it is 10,000.
    Remaining parameters: genome length is 10kb, 20 diploid samples, and
    $r=2.5\times10^{-8}$.
    There are 148 edges in the resulting ARG.
    A single $N_e$ value was estimated for each
    of the three slices of the ARG separated by
    the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. The number of edges per time slice
    (from most recent to oldest) was: 68, 57, and 23. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

>>>>

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
    \caption{Log likelihood-ratio curves for ARGs with various properties.
    The ``Full ARG'' here is the result of an
    msprime simulation of the SMC (100 diploid samples, genome length 1Mb,
    $r=\mu=1.25\times10^{-8}$, $N_e=1\times10^4$).
    The ``RE nodes removed'' ARG is the
    result of simplifying out all of the recombination nodes,
    and the ``Fully simplified'' ARG has all locally unary nodes removed.
    Each of these simulated ARGs additionally has 5\% and 25\% of the
    internal nodes removed to create polytomies.
    We also show the results for ARGs inferred from the original simulation
    data using tsinfer and tsdate
    (see text for details), in which we use the default output of tsinfer
    including unary nodes (tsinfer+tsdate), and in which we fully simplify
    before passing to tsdate (tsinfer+tsdate FS).
    See the text for details on how the likelihood ratio
    values are computed.
    The ``Full ARG'' and ``RE nodes removed'' lines coincide.
    \label{fig:lik-surface}}
\end{figure}

@@

Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
and even on the fully simplified
ARG has a very minor effect on location of the minimum.
Supplemental Fig.~S1 explores more extreme examples in which we remove
25\%, 50\% and 75\% of the internal nodes in these ARGs, and demonstrates 
a remarkable robustness to the loss of the majority of the internal
nodes in these examples. Interestingly, the loss of 75\% of the 
internal nodes  in the fully simplified ARG 
(1841 of 2455, leading  to an average of 27 polytomies per tree)
appears to improve performance, substantially 
reducing the bias away from the true parameter value.

Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from 
the  \revpoint{2}{35}
simulated data.
To do this we first simulated mutations on the full ARG 
at rate $1.25\times10^{-8}$ using msprime, resulting in an 
ARG with 2,989 mutations.
We then used tsinfer and tsdate~\citep{wohns_unified_2022}
to infer a dated ARG,
using default parameters for tsinfer and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the
\cite{li2003modeling} copying process. Information about
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to
attempt to reconstruct the exact subtree-prune-and-regraft
moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well \revpoint{2}{37}
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we
fully simplify the output of tsinfer
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a bias \revpoint{2}{38}
away from the true value of the recombination rate parameter.

\subsection*{Piecewise decomposition over time}
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under the Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices:
    from 0 to 1,000 generations ago $N_e$ is 10,000;
    from 1,000 to 5,000 generations ago it is 5,000;
    and past that it is 10,000.
    Remaining parameters: genome length is 10kb, 20 diploid samples, and
    $r=1.25\times10^{-8}$.
    There are 148 edges in the resulting ARG.
    A single $N_e$ value was estimated for each
    of the three slices of the ARG separated by
    the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. The number of edges per time slice
    (from most recent to oldest) was: 68, 57, and 23. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}
